---
title: "Creating a road elevation map in R"
author: "Jerry Shannon"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
library(osmdata)
library(elevatr)
library(raster)
library(rgdal)
library(tmap)
```

## Downloading OSM data

First I loaded road data for Clarke County, which was obtained directly from the county. This code also includes a method for downloading these data from OpenStreetMap. I did this using the county boundaries, creating a bounding box using st_bbox in sf.  We then use st_intersection from sf to trim to the Clarke County outline. I add a road ID field (`L1`) that will be used to join data further down.

```{r cars}
clarke<-st_read("data/US county 2012_Georgia.geojson", quiet=TRUE) %>%
  filter(GEOID==13059)

bbox_clarke<-st_bbox(clarke)

## Code for OSM data below
# roads<-opq(bbox=bbox_clarke) %>%
#   add_osm_feature(key = 'highway') %>%
#   osmdata_sf()
# 
# roads_sf<-roads$osm_lines %>%
#   st_intersection(clarke) %>%
#   select(osm_id,name)

roads_sf<-st_read("data/RoadCenterline.shp") %>%
  st_transform(4326)

#Add a line number.
roads_sf$L1<-1:nrow(roads_sf)

ggplot(roads_sf) + geom_sf()
```

Since these line segments vary in length, I used `st_segmentize` to add additional points to these lines, with a distance of no more than 15 meters between points. To convert these features from to points, we use st_cast to change the geometry type and then st_coordinates to reduce down to the component points (the vertices of the line segment. The line number created in the previous chunk (variable `L1`) can be used to join point attributes back.

```{r}
roads_point<-roads_sf %>%
  st_transform(32616) %>%
  st_segmentize(15) %>% 
  st_transform(4326) %>%
  st_cast("MULTIPOINT") %>%
  st_coordinates() %>%
  as_tibble()%>%
  st_as_sf(coords=c("X","Y"),crs=4326,remove=FALSE)

ggplot(roads_point %>%
         filter(X > -83.395 & X < -83.385 &
                  Y > 33.945 & Y < 33.95)) + 
  geom_sf()
```

Using the `elevatr` package, I can determine the elevation of each point. First, I used `get_elev_raster` to download a DEM for the study area. According to the [package documentation](https://www.rdocumentation.org/packages/elevatr/versions/0.2.0/topics/get_elev_raster), this DEM is from the Mapzen Terrain Service.

```{r}
dem_area<-get_elev_raster(roads_point,z=13) %>%
  mask(clarke)
raster::plot(dem_area)
```

Then I determine the elevation value (in meters) for each point using the `extract` function in raster. This column of values is connected back to the road points via bind_cols and then the range is calculated for each line segment. Lastly, I join this range value back to the road lines data.

```{r}
roads_point_elev<-raster::extract(dem_area,roads_point) %>%
  as_tibble()

roads_point_elev1<-bind_cols(roads_point,roads_point_elev) %>%
  st_set_geometry(NULL) %>%
  group_by(L1) %>%
  summarise(range=max(value)-min(value))

roads_point_range<-roads_sf %>%
  left_join(roads_point_elev1) 

ggplot(roads_point_range) + 
  geom_sf(aes(color=range))
```

Because lines vary in length, I normalize the change in range by calculating the distance of each line. I use `st_length` for this purpose. I use `bind_cols` to connect this back to the main dataset and then calculate a new variable--angle--that is the ratio of range to distance. There are some upper outliers for this variable due to short or very steep hills, and so I top code it based on the value of 1.5 time the IQR above the third quartile, the standard cutoff for outliers.

```{r}
dist<-st_length(roads_point_range) %>%
  as_tibble() %>%
  mutate(distance=as.numeric(value))

roads_point_all<-roads_point_range %>%
  bind_cols(dist) %>%
  mutate(angle=range/distance)

cutoff<-as.numeric(summary(roads_point_all$angle)[5]+(1.5 * IQR(roads_point_all$angle,na.rm=TRUE)))

roads_point_all<-roads_point_all %>%
  mutate(angle=if_else(angle >= cutoff, cutoff,angle))

tm_shape(dem_area)+
  tm_raster(palette="Greens",alpha=0.4)+
tm_shape(roads_point_all)+
  tm_lines("angle",palette="OrRd",lwd=2,alpha=0.7)+
tm_legend(legend.outside=TRUE)
```
